Parallelization of Fast $\ell_1$-Minimization for Face Recognition

### 2011-04-17 main_allen.tex
Fixed some writing in CPU and GPU alignment implementation sections.

The optimal strategy to implement Algorithm \ref{alg:alignment_alm} on a multicore CPU must take into account
the trade-off between the available memory/cache sizes at multiple hierarchies and their bandwidth based on a particular
CPU architecture. Figure \ref{fig:alignment_mapping_old} shows the the sizes of L2 and L3 caches of the Intel E5530 CPU, 
which is a good example of modern multicore CPU architectures. 
\begin{figure}
\centering
\includegraphics[width=3.4in]{figures/alignment_mapping_old}
\caption{Available L2 and L3 caches on the Intel E5530 multicore CPU architecture with dual CPU configuration.}
\label{fig:alignment_mapping_old}
\end{figure}

As we have calculated in Section \ref{sec:CPU-concurrency}, the L2 cache in
each core is only able to store about 16 images. For image alignment problem,
the number of training images per class is typically larger than 16. Hence, if
\eqref{eq:alignment-linearization} would be solved in parallel on the eight
cores, it would require L3 cache to hold data up to four classes at a time, and
the L2 cache should retrieve subsets of the data frequently during the
iterative process to execute Algorithm \ref{alg:alignment_alm}. Therefore, the
bottleneck in this strategy becomes the bandwidth between the L2 cache and L3
cache. The implementation of some previous works
\cite{WagnerA2009-CVPR,WagnerA2011-PAMI} is indeed based on this mapping
strategy.

In this paper, we show that for CPU parallelism, a more traditional
implementation for pixel-level parallelism in fact outperforms problem-level
parallelism using eight separate cores. In other words, the CPU implementation
performs alignment of a query image against the training images of single or
dual users at a time on the dual processors, depending on the setting of the
compiler. More specifically, in our implementation, most of the operations take
advantage of Intel Math Kernel Library (MKL), a commercial implementation of
the standard Basic Linear Algebra Subprograms (BLAS). For the image resampling
step, we make use of the Intel Integrated Performance Primitives (IPP) library.
Both MKL and IPP are optimized for Intel multicore CPUs, and are able to
automatically utilize both core-level and vector-level concurrency. For other
operations that are not optimized by Intel in-house libraries, we achieve both
vector-level and core-level concurrency through a combination of the Open MP
API \cite{dagum2002openmp} and the automatic vectorization facilities of the
Intel C++ compiler (ICC) \cite{dulong1999overview}. We shall compare the
performance of the two parallelism methods in Section \ref{sec:experiment}.

\subsection{GPU Implementation}
\label{sec:alignment_implementation_gpu}

Similarly, we also consider the trade-off between the available cache sizes and
their bandwidth based on the Nvidia CUDA GPU architecture chosen for the paper.
Figure \ref{fig:alignment_mapping_gpu} illustrates the multicore architecture
and the L1/L2 caches of the Nvidia GTX 480 GPU.
\begin{figure}
\centering
\includegraphics[width=3.4in]{figures/alignment_mapping_gpu}
\caption{Proposed mapping of alignment parallelism onto gpu concurrency}
\label{fig:alignment_mapping_gpu}
\end{figure}

Our proposed GPU implementation for the iterative alignment stage consists of a
single kernel that performs all of the test image alignment problems.  Each of
instance of the alignment problem is assigned its own thread block, and the GPU
hardware schedules as many of these thread blocks to run concurrently as the 
hardware will allow.

@@

In one experiment, we have exploited several possible GPU implementation strategies, which include
solving single alignment problems one at a time using CUBLAS (302\,ms); solving one problem
per SMP using CUDA streams (70\,ms); and solving one problem per SMP using one large kernel call
 (36\,ms). The inferior performance of CUBLAS may be attributed to the relatively small problem
 size. In Section \ref{sec:experiment}, we will further exploit the difference between CPU and GPU
 via synthetic and real image data.

@@

Operations that do not map well onto the BLAS API (mostly combined
vector-vector operations) were implemented as gpu kernels (in C for CUDA).
Even for operations that could have been implemented via multiple BLAS calls,
this was found to have better performance due to reduced bandwidth and kernel
call overhead {\bf Victor, verify or correct this.  I'm not sure, currently asking Mark}.

In order to avoid expensive data transfer across the bus connecting the GPU
card and the CPU motherboard (the PCI express bus), all of the data is
transferred to GPU DRAM once, and all non-trivial tasks are performed on the
GPU on data stored in GPU DRAM.  The sole exception to the above implementation
strategy is that the pseudoinverse of A is still computed on the CPU via the
LAPACK implementation that is part of Intel's MKL. This operation need only be
performed (and results uploaded to GPU DRAM) once before GPU computation
starts. {\bf we should prove justification for A pseudoinverse being cheap,
either here or perhaps later in a pie chart}

>>>>

Optimal implementation of Algorithm \ref{alg:alignment_alm} on a multicore CPU must take 
into account the properties of the cache hierarchy. In general, cache that is 
closer to the core (i.e. L1 cache) has higher bandwidth, but smaller size compared to cache
that is farther from the core (L3 cache).  For reference, Figure \ref{fig:caches} 
shows the the sizes of L2 and L3 caches of the Intel E5530 CPU, 
which is a representative example of a modern multicore CPU.

For the E5530, the L2 cache in each core is only able to store about 16 images.
For image alignment problem, the number of training images per class is
typically larger than 16.  Therefore, if eight instances of
\eqref{eq:alignment-linearization} were be executed in parallel on the eight
cores, there would be an insufficient amount of L1 or L2 cache to hold $A$ and
$A^\dagger$ during the iterative process to execute Algorithm
\ref{alg:alignment_alm}. Therefore, the bottleneck in this strategy becomes the
bandwidth between the L2 cache and L3 cache. The implementation of some
previous works \cite{WagnerA2009-CVPR,WagnerA2011-PAMI} is indeed based on this
mapping strategy.

In this paper, we show that for CPU parallelism of face alignment, a higher
performance implementation can be achieved by {\em not} taking advantage of
problem-level parallelism at all, and instead map the pixel-level parallelism
onto both the core-level and the sse-level concurrency provided by the cpu.  In
other words, the proposed CPU implementation performs alignment of a query
image against the training images of a single user at a time using all cores on
both processors.
This mapping of the problem is illustrated in Figure \ref{fig:alignment_mapping_cpu}.

More specifically, in our implementation, most of the operations take
advantage of Intel Math Kernel Library (MKL), a commercial implementation of
the standard Basic Linear Algebra Subprograms (BLAS). For the image resampling
step, we make use of the Intel Integrated Performance Primitives (IPP) library.
Both MKL and IPP are optimized for Intel multicore CPUs, and are able to
automatically utilize both core-level and vector-level concurrency. For other
operations that are not optimized by Intel in-house libraries, we achieve both
vector-level and core-level concurrency through a combination of the Open MP
API \cite{dagum2002openmp} and the automatic vectorization facilities of the
Intel C++ compiler (ICC) \cite{dulong1999overview}. We shall compare the
performance of the two parallelism methods in Section \ref{sec:experiment}.

\subsection{GPU Implementation}
\label{sec:alignment_implementation_gpu}
While on the CPU, algorithm performance is highly dependent on effective
cache usage, on the GPU cache is relatively unimportant for ALM based
$\ell_1$-min.  As can be seen in Figure \ref{fig:caches}, the GPU has 
a very small amount of cache compared to the CPU.  While it might be possible
to fit a single instance of the alignment problem (with a slightly reduced problem size)
into L2 cache, this would not be an efficient use of the GPU's resources.
The strength of the GPU's memory architecture for our purposes it's ability to sustain a very
high bandwidth to DRAM.  A higher overall bandwidth can be achieved if 
many memory transactions are performed simultaneously, than can be achieved
with a large number of smaller transactions that are just large enough to fill
the L2 cache.

Therefore, whereas on the CPU we advocate performing alignment for one training
class at a time, on the GPU we advocate performing alignment for many classes
concurrently.  \footnote{In CUDA terminology, Our proposed alignment stage
implementation consists of a single kernel that performs alignment for all
subject classes.  Each instance of the alignment problem is assigned its own
thread block, and the GPU hardware schedules as many thread blocks to run
concurrently as the hardware will allow.} The number of subject classes that
actually get aligned concurrently is chosen by the gpu hardware. In our
experiments the hardware is able to schedule 4-5 classes per SMP to run
concurrently. This mapping of the problem-level parallelism onto the SMP-level
concurrency is illustrated in Figure \ref{fig:alignment_mapping_gpu}.

\begin{figure}
\centering
\includegraphics[width=3.4in]{figures/alignment_mapping_gpu}
\caption{Proposed mapping of alignment parallelism onto gpu concurrency}
\label{fig:alignment_mapping_gpu}
\end{figure}

@@

In one experiment, we have exploited several possible GPU implementation
strategies, which include solving single alignment problems one at a time using
CUBLAS (302\,ms); solving one problem per SMP using CUDA streams (70\,ms); and
solving one problem per SMP using one large kernel call (36\,ms). The inferior
performance of CUBLAS may be attributed to the relatively small problem size.
In general, the libraries provided by Nvidia are optimized for problem sizes
that are significantly larger than our individual alignment problems.  In
Section \ref{sec:experiment}, we show the performance of our CPU and GPU
implementations running on synthetic and real image data.

@@

Operations that do not map well onto the BLAS API (mostly combined
vector-vector operations) were implemented as gpu kernels (in C for CUDA).
Even for some operations that could have been implemented via multiple BLAS
calls, performance improvements were achieved by combining multiple
vector-vector operations into a single kernel, due to reduced bandwidth and
kernel call overhead.

In order to avoid expensive data transfer across the bus connecting the GPU
card and the CPU motherboard (the PCI express bus), all of the data is
transferred to GPU DRAM once, and all non-trivial tasks are performed on the
GPU on data stored in GPU DRAM.  The sole exception to the above implementation
strategy is that the pseudoinverse of A is still computed on the CPU via the
LAPACK implementation that is part of Intel's MKL. This operation need only be
performed (and results uploaded to GPU DRAM) once before GPU computation
starts. Since the $A$ is very tall, it is only necessary to invert a matrix
of size $n \times n$, which executes very quickly.
